A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-19 tex/body.tex
looks better

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C},\, \const{D},\, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
z \not\teq v
\}$.
After applying the calculus to saturation on $\Ec$,
the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( {\Varec{z}} ) &
\Val \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( {\Varec{y}} ) &
\Val \ec{y} & {}\,=\,{} & {\Varec{y}} \\
\Val \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( {\Varec{z}} ) &
\Val \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( {\Varec{z}} ) ) &
\Val \ec{z} & = & {\Varec{z}}
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, specifically ${\Varec{y}}$ and ${\Varec{z}}$.
Modulo $\alpha$-equivalence, $\smash{\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )} =
%\begin{array}{c}
\{
\MU a.\: \const{C}( a ),\:
\MU a.\: \const{D}( a ),\allowbreak\:
\MU a.\: \const{C}( \const{D}( a )),\allowbreak\:
\const{C}( \MU a.\: \const{D}( a ))
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that evaluates variables in $\Ec$ based on the mapping $\Val$:
$z$ is interpreted as $\Val \ec{z}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{z}$ that is $\alpha$-equivalent to a value in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$ for ${\Varec{z}}$.
After the substitution, $\Val \ec{v} = \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{D}( {\Varec{z}} ) )$,
which has normal form $\MU {\Varec{v}}.\; \const{D}( {\Varec{v}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$.
However, this contradicts the disequality $z \not\teq v$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{z}}$ is fresh, %not $\alpha$-equivalent to any term in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$ or
$\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$
to ${\Varec{z}}$.
Legal substitutions for ${\Varec{z}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$ to $\Varec{z}$, we have
$\Val \ec{u} = \MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) ) )$,
with normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( {\Varec{u}} ) )$.
\xend
\end{examplex}

@@

\begin{lemma}
\label{lem:interpretation-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal interpretation $\J$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x\typ{\tau}}$ is $\alpha$-equivalent to a value in $\interp{\J}{\tau}$.
\item[\rm (2)]
$\ValC \ec{x} = \expand{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$.
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{rep}
\begin{proof}
To show (1), we first show that $\ValC$ contains no free variables.
Assume by contradiction that $\ValC$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ of type $\tau$.
This is a contradiction, since %our %model
completion can assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$,
and thus %our %model
completion can assign a value not occurring in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ to ${\Varec{y}}$.
By construction, $\ValC \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, the construction of $\ValC$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\ValC \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\interp{\J}{\tau}$.

We first show that (2) and (3) hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_{\negvthinspace j}( \vec t )$ and $\const{C}_{\negvthinspace\jPrime}( \vec u )$ in $\ec{z}$
are such that $j = \jPrime$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply to $\Ec$.
Thus,
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_{\negvthinspace j}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_{\negvthinspace j}( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $t$ that are bound in $\Val \ec{y}$ are interpreted in
the expansion of $\expand{\Val \ec{y}}{ t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. %\rem{TODO: more detail}
Thus, we have
$\expand{\Val \ec{y}}{ t } = \expand{\Val \ec{x}}{ \Val \ec{x} } = \Val \ec{x}$, and thus (2) holds for $\Val$.
Property (3) holds for $\Val$ since \rn{Unique} does not apply.

We now show that $\Val = \nf{\Val}$.
Assume by contradition $\Val \ec{x} \neq \nf{\Val \ec{x}}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to the construction of $\Val$,
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for some $i$ % \in [n]$.
Since $\Val \ec{x}$ is a minimal, it
contains a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\expand{\Val \ec{x}}{ t_i } \aequiv \expand{\Val \ec{x} }{ u_i }$ for some $i$. % \in [n]$.
Due to the construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for some $i$. % \in [n]$.
Since \rn{Cong} does not apply to $\Ec$
we have $\ec{w_{\negvthinspace j}}$ and $\ec{z_{\negvthinspace j}}$ are distinct for some $j$.
By (2), $\expand{\Val \ec{x}}{ t_{\negvthinspace j} } = \Val \ec{z_{\negvthinspace j}}$
and $\expand{\Val \ec{x}}{ u_{\negvthinspace j} } = \Val \ec{w_{\negvthinspace j}}$,
which are not $\alpha$-equivalent by (3), contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a self-similar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{\Val}$, and (2) and (3) hold for $\nf{\Val}$.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\interp{\J}{t}$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \interp{\J}{u_1}, \ldots, \interp{\J}{u_n} ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \ValC \ec{ u_1 }, \ldots, \ValC \ec{ u_n } )}$.
Call this term $t_u$.
%By Lemma~\ref{lem:interpretation-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\expand{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by the construction of $\ValC$ we have $\ValC \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
where
$\muvar(w_i) = {\Varec{u_i}}$ for each $i$.  Thus by Lemma~\ref{lem:interpretation-completion}\PART{2},
$\expand{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$.
For each $i$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
\begin{rep}By Lemma~\ref{lem:mu-norm-arg},\end{rep}%
\begin{conf}Clearly,\end{conf} %% TODO: Fix! ;)
$\expand{t_u}{u_\iPrime} \vsim \ValC \ec{u_i}$.
Thus $\expand{t_u}{u_\iPrime} \vsim \expand{\ValC \ec{t}}{ w_i }$. % for each $i \in [n]$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\interp{\J}{t} \vsim t_u \vsim \ValC \ec{t}$, and
we have $\interp{\J}{t} \vsim \ValC \ec{t}$.

If $t$ is a selector term %of the form
$\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\smash{\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)}$ for some $\jPrime$.
%By Lemma~\ref{lem:interpretation-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\expand{\ValC \ec{u} }{ w_i } = \ValC \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$,
by construction $\ValC$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$
for each $i$ and thus by Lemma~\ref{lem:interpretation-completion}\PART{2},
$\expand{\ValC \ec{u}}{ w_i } = \ValC \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$.
If $j = \jPrime$, then $\interp{\J}{t}$ is $\alpha$-equivalent with $\expand{\ValC \ec{u}}{w_k}$, which is equal to $\ValC \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\ValC \ec{t}$.
If $j \neq \jPrime$, since \rn{Cong} does not apply,
any term of the form $\smash{\const s^k_{\negvthinspace j}( u' )}$ not occurring in $\ec{t}$ 
is such that $u$ and $u'$ are in distinct equivalence classes.
Due to the induction hypothesis and Lemma~\ref{lem:interpretation-completion}\PART{3}, $\interp{\J}{u} \neq \interp{\J}{u'}$ for all such $u$, $u'$.
Thus, we may interpret $\smash{\interp{\J}{\const s^k_{\negvthinspace j}}( \interp{\J}{u} )}$ as the value in $\interp{\J}{\tau}$ that is $\alpha$-equivalent with $\ValC \ec{ t }$.

We now show that all disequalities in $\Fc$ are satisfied by $\J$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\interp{\J}{t_1} \vsim \ValC \ec{t_1}$ and $\interp{\J}{t_2} \vsim \ValC \ec{t_2}$,
by Lemma~\ref{lem:interpretation-completion}\PART{3}, $\interp{\J}{t_1} \neq \interp{\J}{t_2}$, and thus $\J \models t_1 \tneq t_2$.

@@

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
%
Discriminators are part of the signature instead
of being abbreviations.
We have found that this leads to better performance since it introduces terms less eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, we immediately add
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}

>>>>

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C},\, \const{D},\, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
u \teq \const{C}( z ),\;
v \teq \const{D}( z ),\;
w \teq \const{E}( y ),\;
x \teq \const{C}( v ),\;
z \not\teq v
\}$.
After applying the calculus to saturation on $\Ec$,
the mapping $\Val$ is as follows:
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
%\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l}
\[\begin{array}{r@{}c@{}r@{}l@{\qquad}r@{}c@{}r@{}l@{\qquad}r@{}c@{}l}
\Val \ec{u} & {}\,=\,{} & \MU {\Varec{u}}.\; & \const{C}( {\Varec{z}} ) &
\Val \ec{w} & = & \MU {\Varec{w}}.\; & \const{E}( {\Varec{y}} ) &
\Val \ec{y} & {}\,=\,{} & {\Varec{y}} \\
\Val \ec{v} & = & \MU {\Varec{v}}.\; & \const{D}( {\Varec{z}} ) &
\Val \ec{x} & {}\,=\,{} & \MU {\Varec{x}}.\; & \const{C}( \MU {\Varec{v}}.\; \const{D}( {\Varec{z}} ) ) &
\Val \ec{z} & = & {\Varec{z}}
\end{array}\]
%
To construct a completion $\ValC$, we must choose values for the free
variables of $\Val$, specifically ${\Varec{y}}$ and ${\Varec{z}}$.
Modulo $\alpha$-equivalence, $\smash{\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )} =
%\begin{array}{c}
\{
\MU a.\: \const{C}( a ),\:
\MU a.\: \const{D}( a ),\allowbreak\:
\MU a.\: \const{C}( \const{D}( a )),\allowbreak\:
\const{C}( \MU a.\: \const{D}( a ))
\}
%\end{array}
$.
Consider a normal interpretation $\J$ that evaluates variables in $\Ec$ based on the mapping $\Val$:
$z$ is interpreted as $\Val \ec{z}$, $y$ as $\Val \ec{y}$, and so on.
Assigning a value for $\Val \ec{z}$ that is $\alpha$-equivalent to a value in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\Ec$ to be falsified by $\J$.
For example, assign $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$ for ${\Varec{z}}$.
After the substitution, $\Val \ec{v} = \MU {\Varec{v}}.\; \const{D}( \MU {\Varec{z}}.\; \const{D}( {\Varec{z}} ) )$,
which has normal form $\MU {\Varec{v}}.\; \const{D}( {\Varec{v}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{z}}.\; \const{D}( {\Varec{z}} )$.
However, this contradicts the disequality $z \not\teq v$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{z}}$ is fresh, %not $\alpha$-equivalent to any term in $\Terms^{\smash{\J}}_{{\Varec{z}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign a value such as
$\MU {\Varec{z}}.\; \const{E}( {\Varec{z}} )$ or
$\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$
to ${\Varec{z}}$.
\begin{rep}Legal substitutions for ${\Varec{z}}$ may cause the range of $\Val$ to contain abnormal terms.
For example, after assigning $\MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) )$ to $\Varec{z}$, we have
$\Val \ec{u} = \MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( \const{C}( {\Varec{z}} ) ) )$,
with normal form
$\MU {\Varec{u}}.\; \const{C}( \MU {\Varec{z}}.\; \const{D}( {\Varec{u}} ) )$.\end{rep}
\xend
\end{examplex}

@@

\begin{lemma}
\label{lem:interpretation-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal interpretation $\J$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x\typ{\tau}}$ is $\alpha$-equivalent to a value in $\interp{\J}{\tau}$.
\item[\rm (2)]
$\ValC \ec{x} = \expand{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$.
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{rep}
\begin{proof}
To show (1), we first show that $\ValC$ contains no free variables.
Assume by contradiction that $\ValC$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Single} does not apply, % to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ of type $\tau$.
This is a contradiction, since %our %model
completion can assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$,
and thus %our %model
completion can assign a value not occurring in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ to ${\Varec{y}}$.
By construction, $\ValC \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, the construction of $\ValC$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\ValC \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\interp{\J}{\tau}$.

We first show that (2) and (3) hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_{\negvthinspace j}( \vec t )$ and $\const{C}_{\negvthinspace\jPrime}( \vec u )$ in $\ec{z}$
are such that $j = \jPrime$ since \rn{Clash} does not apply, % to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply. % to $\Ec$.
Thus,
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_{\negvthinspace j}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_{\negvthinspace j}( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $t$ that are bound in $\Val \ec{y}$ are interpreted in
the expansion of $\expand{\Val \ec{y}}{ t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. %\rem{TODO: more detail}
Thus, we have
$\expand{\Val \ec{y}}{ t } = \expand{\Val \ec{x}}{ \Val \ec{x} } = \Val \ec{x}$, and thus (2) holds for $\Val$.
Property (3) holds for $\Val$ since \rn{Unique} does not apply.

We now show that $\Val = \nf{\Val}$.
Assume by contradition $\Val \ec{x} \neq \nf{\Val \ec{x}}$ for some $\Val \ec{x}$ of minimal size.
We have that $\Val \ec{x}$ is of the form $\MU {\Varec{x}}.\; \const{C}( t_1, \ldots, t_n )$.
Due to the construction of $\Val$,
we know $\ec{x}$ contains a constructor $\const{C}( z_1, \ldots, z_n )$ and $\muvar( t_i ) = z_i$ for some $i$ % \in [n]$.
Since $\Val \ec{x}$ is a minimal, it
contains a subterm of the form $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$
where $\expand{\Val \ec{x}}{ t_i } \aequiv \expand{\Val \ec{x} }{ u_i }$ for some $i$. % \in [n]$.
Due to the construction of $\Val$, $\ec{y}$ contains a constructor $\const{C}( w_1, \ldots, w_n )$ and $\muvar( u_i ) = w_i$ for some $i$. % \in [n]$.
Since \rn{Cong} does not apply, % to $\Ec$
we have $\ec{w_{\negvthinspace j}}$ and $\ec{z_{\negvthinspace j}}$ are distinct for some $j$.
By (2), $\expand{\Val \ec{x}}{ t_{\negvthinspace j} } = \Val \ec{z_{\negvthinspace j}}$
and $\expand{\Val \ec{x}}{ u_{\negvthinspace j} } = \Val \ec{w_{\negvthinspace j}}$,
which are not $\alpha$-equivalent by (3), contradicting the fact that $\MU {\Varec{y}}.\; \const{C}( u_1, \ldots, u_n )$ is a self-similar subterm of $\Val \ec{x}$.
Thus, $\Val = \nf{\Val}$, and (2) and (3) hold for $\nf{\Val}$.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\interp{\J}{t}$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \interp{\J}{u_1}, \ldots, \interp{\J}{u_n} ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \ValC \ec{ u_1 }, \ldots, \ValC \ec{ u_n } )}$.
Call this term $t_u$.
%By Lemma~\ref{lem:interpretation-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\expand{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply, % to $\Fc$,
by the construction of $\ValC$ we have $\ValC \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
where
$\muvar(w_i) = {\Varec{u_i}}$ for each $i$.  Thus by Lemma~\ref{lem:interpretation-completion}\PART{2},
$\expand{\ValC \ec{t}}{w_i} = \ValC \ec{u_i}$.
For each $i$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
\begin{rep}By Lemma~\ref{lem:mu-norm-arg},\end{rep}%
\begin{conf}Clearly,\end{conf} %% TODO: Fix! ;)
$\expand{t_u}{u_\iPrime} \vsim \ValC \ec{u_i}$.
Thus $\expand{t_u}{u_\iPrime} \vsim \expand{\ValC \ec{t}}{ w_i }$. % for each $i \in [n]$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\interp{\J}{t} \vsim t_u \vsim \ValC \ec{t}$, and
we have $\interp{\J}{t} \vsim \ValC \ec{t}$.

If $t$ is a selector term %of the form
$\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\smash{\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)}$ for some $\jPrime$.
%By Lemma~\ref{lem:interpretation-completion}\PART{2} and due to our construction of $\ValC$, we have $\ValC \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\expand{\ValC \ec{u} }{ w_i } = \ValC \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i \in [n]$.
Since \rn{Inject} and \rn{Clash} do not apply, % to $\Fc$,
by construction $\ValC$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$
for each $i$ and thus by Lemma~\ref{lem:interpretation-completion}\PART{2},
$\expand{\ValC \ec{u}}{ w_i } = \ValC \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$.
If $j = \jPrime$, then $\interp{\J}{t}$ is $\alpha$-equivalent with $\expand{\ValC \ec{u}}{w_k}$, which is equal to $\ValC \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}
= \ValC \ec{t}$.
If $j \neq \jPrime$, since \rn{Cong} does not apply,
any term of the form $\smash{\const s^k_{\negvthinspace j}( u' )}$ not occurring in $\ec{t}$ 
is such that $[u] \not= [u']$.
By the induction hypothesis and Lemma~\ref{lem:interpretation-completion}\PART{3}, $\interp{\J}{u} \neq \interp{\J}{u'}$ for all such $u$, $u'$.
Thus, we may interpret $\smash{\interp{\J}{\const s^k_{\negvthinspace j}}( \interp{\J}{u} )}$ as the value in $\interp{\J}{\tau}$ that is $\alpha$-equivalent with $\ValC \ec{ t }$.

We now show that all disequalities in $\Fc$ are satisfied by $\J$.
Assume $t \tneq u \in \Fc$.
Since \rn{Conflict} does not apply, $t \teq u \not\in \Fc$ and thus $\ec{t}$ and $\ec{u}$ are distinct.
Since $\interp{\J}{t} \vsim \ValC \ec{t}$ and $\interp{\J}{u} \vsim \ValC \ec{u}$,
by Lemma~\ref{lem:interpretation-completion}\PART{3}, $\interp{\J}{t} \neq \interp{\J}{u}$, and thus $\J \models t \tneq u$.

@@

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
\begin{rep}We briefly mention the main ones. \end{rep}%
Discriminators are part of the signature and not abbreviations.
This requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to datatypes and codatatypes.
This approach often leads to better performance because it introduces terms
less eagerly to $\tEc$.
Selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, the solver directly adds
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}
